State of the Practice for Lattice Boltzmann Method Software

### 2022-02-10 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Addition of some notes for planning the revisions of the next sections

During our interviews, documentation was noted as playing a significant role in
the development process, specifically with on-boarding new developers. A goal of
documentation is to lower the entry barrier for these new contributors. The
documentation provides information on how to get started, orients the user to
artifacts and the source code, and explains how the system works, including the
so-called simulation engine and interface.  This emphasis on documentation,
especially for new developers, is echoed in research software guidelines.
Multiple guidelines recommend a document explaining how to contribute to a
project, often named CONTRIBUTING \citep{Yo2021, BrettEtAl2021, WilsonEtAl2016,
ThielEtAl2020, vanGompelEtAl2016, OrvizEtAl2017, FLOSS2022, JimenezEtAl2017}.
Tutorials \citep{ThielEtAl2020}, trouble shooting guides \citep{OrvizEtAl2017,
SSI2022} and quick start examples \citep{ThielEtAl2020, vanGompelEtAl2016} are
also recommended.  \citep{SmithAndRoscoe2018} suggests including instructions
specifically for on-boarding new developers. For open source software in
general, \citep{Fogel2005} recommends providing tutorial style examples,
developer guidelines, demos and screenshots.

@@

Answers \rqref{RQ_PainPoints} and \rqref{RQ_ComparePainPoints}.  This subsection
answers the research question: What are the pain points for developers working
on research software projects? What aspects of the existing processes,
methodologies and tools do they consider as potentially needing improvement? How
should processes, methodologies and tools be changed to improve software
development and software quality? Developers were asked to comment on obstacles
in their development process, obstacles encountered by users, and potential
future obstacles.

@@

A developer of pyLBM noted that their small development team has a lack of time
to implement new features. Small development teams are common for LBM software
packages. Team members are almost always part of the same institute or already
know each other from other projects. External contributions are rare despite
many of the projects accepting them. \wss{Other projects cite the same point
point - Wiese paper}  Aside from on-boarding new developers, time constraints
could be mitigated by increasing developer efficiency, which could be addressed
in several ways, including by improving the quality of documentation, or
incorporating automatic code generation. \wss{Potential ways to address the pain
points might go into a separate section?  Maybe propose the strategy, like code
generation, and then list the pain points that are addressed?}

@@

Interviewees commented that documentation is important and that its quality
could be improved. As already noted, there is often no time or funding for
maintaining quality documentation for software that is rarely used outside of
the development team. Furthermore, the documentation generally only provides a
shallow overview of the underlying CFD theory. Users would be well advised to
already be familiar with these topics, or they should spend significant time
referencing theory resources. The documentation instead generally focuses on
explaining how to use the software. It is of course not feasible for package
documentation to address the underlying physics topics in detail, so it is
advised that the package documentation links to resources that better explain
the underlying theory. Sometimes, frequently asked questions about the
underlying theory are answered in the documentation. OpenLB has an artifact for
such questions.  \wss{Weise mentions documentation.}

@@

Although not always visible from the publicly available artifacts, interviews
with developers revealed that most projects have a defined process for accepting
contributions from team members. The packages rarely get contributions from
outside developers, but the process would be similar as for the aforementioned
group.  In general, contributions are made through GitHub, and are then reviewed
and pulled by lead developers, often with consultation with a group of core
developers depending on the organizational model. Continuous integration is part
of the process for some packages. \wss{Recommendation to increase.  Successful
projects use it, a reasonable next step for projects that have not yet
incorporated CI.}

Some developers noted that their software package does not have well defined
contributing guide in the repository, but it might be a good idea to add one in
the near future. They would be happy to see contribution from outside of their
organization, but currently this does not happen.  \wss{A comment was made at
the recent DOE workshop that you get new users through documentation and
tutorials.  Maybe highlight the example of Musubi?  Should trace back to this
for the recommendations part.}

@@

Software packages were not quantitatively measured for modifiability. In this
project we asked developers to comment on modifiability when we interviewed
them. Specifically, we asked if ease of future changes to the system, modules,
and code blocks was considered when designing the software. We also asked if any
measures had been taken to ensure the ease of future changes. All of the
developers that were interviewed noted that the ease of future changes was
considered and that measures to ensure it had been taken, including requiring
the separation of software components in the source code architecture.  \wss{A
good approach for other developers to adopt - part of recommendations.  Trace
back to this comment.}  A high degree of code modularity and abstraction was
noted by developers as a measure to ensure the ease of future changes. This can
be ensured by separating components and hiding information behind well defined
interfaces. The developer of ESPResSo also noted that some of the code base was
transitioned from C to C++, which could ease modifiability of that software
package. The developer of TCLB noted that their software package was designed to
allow for the addition of some LBM features, but changes to major aspects of the
system would be difficult. For example, ``implementing a new model will be an
easy contribution'', but changes to the ``Cartesian mesh...will be a
nightmare''. Furthermore, the package was designed with flexible data structures
and storage in mind.

>>>>

During our interviews, documentation was noted as playing a significant role in
the development process, specifically with on-boarding new developers. A goal of
documentation is to lower the entry barrier for these new contributors. The
documentation provides information on how to get started, orients the user to
artifacts and the source code, and explains how the system works, including the
so-called simulation engine and interface.  This emphasis on documentation,
especially for new developers, is echoed in research software guidelines.
Multiple guidelines recommend a document explaining how to contribute to a
project, often named CONTRIBUTING \citep{Yo2021, BrettEtAl2021, WilsonEtAl2016,
ThielEtAl2020, vanGompelEtAl2016, OrvizEtAl2017, FLOSS2022, JimenezEtAl2017}.
Tutorials \citep{ThielEtAl2020}, trouble shooting guides \citep{OrvizEtAl2017,
SSI2022} and quick start examples \citep{ThielEtAl2020, vanGompelEtAl2016} are
also recommended.  \citep{SmithAndRoscoe2018} suggests including instructions
specifically for on-boarding new developers. For open source software in general
(not just research software), \citep{Fogel2005} recommends providing tutorial
style examples, developer guidelines, demos and screenshots.

@@

This section aims to answer the research questions: What are the pain points for
developers working on research software projects (\rqref{RQ_PainPoints})?; and,
How do the pain points of developers from LBM compare to the pain points for
research software in general (\rqref{RQ_ComparePainPoints})?

Roadmap.  Go through pain points, link to existing literature, then section on
pain points not mentioned.  Ways to address pain points in the next section

Maybe just do a table of pain points?  and then contrast the pain points with
the literature.  The mitigation strategies should be elsewhere.  Some of the
mitigation strategies have already been mentioned, like documentation.

Likely possible to combine pain points and developer concerns.  Emphasizes what
the pain points are, and how they compare to other studies.  Don't propose
solutions in this section.  Say that all of the information comes from the
interviews (as mentioned at the start of the next section).

@@

A developer of pyLBM noted that their small development team has a lack of time
to implement new features. Small development teams are common for LBM software
packages. Team members are almost always part of the same institute or already
know each other from other projects. External contributions are rare despite
many of the projects accepting them. \wss{Other projects cite the same pain
point - Wiese paper}  Aside from on-boarding new developers, time constraints
could be mitigated by increasing developer efficiency, which could be addressed
in several ways, including by improving the quality of documentation, or
incorporating automatic code generation. \wss{Potential ways to address the pain
points might go into a separate section?  Maybe propose the strategy, like code
generation, and then list the pain points that are addressed?}

@@

Interviewees commented that documentation is important and that its quality
could be improved. As already noted, there is often no time or funding for
maintaining quality documentation for software that is rarely used outside of
the development team. Furthermore, the documentation generally only provides a
shallow overview of the underlying CFD theory. Users would be well advised to
already be familiar with these topics, or they should spend significant time
referencing theory resources. The documentation instead generally focuses on
explaining how to use the software. It is of course not feasible for package
documentation to address the underlying physics topics in detail, so it is
advised that the package documentation links to resources that better explain
the underlying theory. Sometimes, frequently asked questions about the
underlying theory are answered in the documentation. OpenLB has an artifact for
such questions.  \wss{Weise mentions documentation.}  \wss{Partly solve the
problem by documentation user characteristics, to explicitly state what the user
is assumed to already know.}

@@

Although not always visible from the publicly available artifacts, interviews
with developers revealed that most projects have a defined process for accepting
contributions from team members. The packages rarely get contributions from
outside developers, but the process would be similar as for the aforementioned
group.  In general, contributions are made through GitHub, and are then reviewed
and pulled by lead developers, often with consultation with a group of core
developers depending on the organizational model. Continuous integration is part
of the process for some packages. \wss{Recommendation to increase.  Successful
projects use it, a reasonable next step for projects that have not yet
incorporated CI.}  \wss{There is definitely a great deal of repetition.}

Some developers noted that their software package does not have well defined
contributing guide in the repository, but it might be a good idea to add one in
the near future. They would be happy to see contribution from outside of their
organization, but currently this does not happen.  \wss{A comment was made at
the recent DOE workshop that you get new users through documentation and
tutorials.  Maybe highlight the example of Musubi?  Should trace back to this
for the recommendations part.  \citep{Fogel2005} gives advice on having a
successful project, also the web article (if needed).}

@@

Software packages were not quantitatively measured for modifiability. In this
project we asked developers to comment on modifiability when we interviewed
them. Specifically, we asked if ease of future changes to the system, modules,
and code blocks was considered when designing the software. We also asked if any
measures had been taken to ensure the ease of future changes. All of the
developers that were interviewed noted that the ease of future changes was
considered and that measures to ensure it had been taken, including requiring
the separation of software components in the source code architecture.  \wss{A
good approach for other developers to adopt - part of recommendations.  Trace
back to this comment.}  A high degree of code modularity and abstraction was
noted by developers as a measure to ensure the ease of future changes. This can
be ensured by separating components and hiding information behind well defined
interfaces. The developer of ESPResSo also noted that some of the code base was
transitioned from C to C++, which could ease modifiability of that software
package. The developer of TCLB noted that their software package was designed to
allow for the addition of some LBM features, but changes to major aspects of the
system would be difficult. For example, ``implementing a new model will be an
easy contribution'', but changes to the ``Cartesian mesh...will be a
nightmare''. Furthermore, the package was designed with flexible data structures
and storage in mind.  \wss{example suggests identifying likely changes.}
